Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Feb 9 09:38:49 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/dc55a82ecaa6c7579c80e824377410a6b61dc53f}{dc55a82ecaa6c7579c80e824377410a6b61dc53f}}

### 2017-03-14 sftrends_manu.Rnw
to STICS
minor yongshan edits

\author{Marcus W. Beck}
\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
\email{beck.marcus@epa.gov}
\phone{+1 (850)9342480}
\fax{+1 (850)9342401}
\author{David Senn}
\author{Thomas Jabusch}
\author{Phil Trowbridge}
\affiliation{San Francisco Estuary Institute, Richmond, CA}

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people and includes irrigation for 18 thousand km$^2$ of agricultural land.  Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.

@@

Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes., particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.

>>>>

\author{Marcus W. Beck}
\affiliation{USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, Gulf Breeze, FL}
\email{beck.marcus@epa.gov}
\phone{+1 (850)9342480}
\fax{+1 (850)9342401}
\author{David B. Senn}
\author{Thomas W. Jabusch}
\author{Phil Trowbridge}
\affiliation{San Francisco Estuary Institute, Richmond, CA}

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides water to over 25 million people and irrigation for 18 thousand km$^2$ of agricultural land.  Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.

@@

Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.

### 2017-03-14 sftrends_manu.tex
minor yongshan edits

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Sun Mar 5 13:06:41 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/bf69bdbb5eef9d3a08d415c53dd23ab29126beda}{bf69bdbb5eef9d3a08d415c53dd23ab29126beda}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

>>>>

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Thu Mar 9 13:51:29 2017 -0600, \href{https://github.com/fawda123/sftrends_manu/commit/3a2258283f79a8ff3710c96c42f090e653e0a955}{3a2258283f79a8ff3710c96c42f090e653e0a955}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers
